FAIR Theory: Applying Open Science Principles to the Construction and Iterative Improvement of Scientific Theories

### 2024-12-16 manuscript.Rmd
final thoughts

## Some other stuff

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code.
Using version control systems, like Git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
Git also facilitates diffuse and adversarial collaboration,
as independent researchers can create independent versions of existing theories through "forking", or suggest modifications to existing theories via "pull requests".
In sum, version control using Git enables systematic, collaborative, and transparent theory development,
enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].

@@

Making Theories FAIR Accelerates Scientific Progress
Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
As several authors have taken inspiration from the work by De Groot,
we compare our interpretation of the original theory to the interpretation of others.
Originally, the theory has the following structure:

@@

@@

One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR Theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of

>>>>

## Additional considerations

We can take inspiration from the field of computer science for well-established processes for iteratively improving information artefacts,  like computer code (which we also have successfully applied in the domain of reproducible research findings, see XXX). <!--AB cite WORCS and Aarons tutorial on how to do the same thing again..-->
Using version control systems, like git, would enhance the reusability of FAIR theory by thoroughly documenting every modification in a traceable and reversible manner.
Git also facilitates diffuse and adversarial collaboration,
as independent researchers can create independent versions of existing theories through "forking", or suggest modifications to existing theories via "pull requests".
In sum, version control using Git enables systematic, collaborative, and transparent theory development,
enables studying the provenance of a theory and investigating how well different iterations of the theory explain empirical evidence [@vanlissaUsingEndpointsCheck2023].

@@

### Making Theories FAIR Accelerates Scientific Progress
Adopting the FAIR principles for theories can address key challenges in the current research landscape, where theories often remain isolated and underutilized. By making theories findable, accessible, interoperable, and reusable, researchers can ensure that their work is grounded in a shared, transparent, and cumulative body of knowledge. This approach enhances scholarly communication, allowing for greater scrutiny, replication, and collaboration across disciplines, ultimately leading to faster, more reliable, and more impactful scientific progress.

@@

In this example, we represent the empirical cycle - a theory of cumulative knowledge production through scientific research - as FAIR theory.
As several authors have taken inspiration from the work by De Groot  [@degrootMethodologieGrondslagenVan1961],  
we compare our interpretation of the original theory to the interpretation of others.
Originally, the theory has the following structure:

@@

<!--AB TODO: Also add images of the graph syntax - not everyone will be able to read this in syntax and see the graph in their minds-->

@@

One remaining issue that intersects with FAIR Theory is the measurement and operationalization of psychological constructs.
Aside from the aforementioned "theory crisis", there has been talk of a "measurement crisis":  <!--AB very important point that I was already missing in the introduction; seems we also need FAIR measurements :) -->
it is not always clear how theoretical constructs are operationalized, and many existing instruments have poor psychometric properties [REF].
Additionally, the "jingle-jangle" fallacy is prevalent in the social sciences:
the same term is often used for distinct constructs, and conversely, different terms are used to refer to the same construct.
FAIR Theory can help address the measurement crisis:
since theories can reference other theories and resources, it is possible to extend a structural theory with a theory of
